Towards an Ecological Trait-data Standard

### 2017-11-09 manuscript/manuscript.Rmd
major revision manuscript

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

1.	The assessment and analysis of functional traits has been established in all domains of ecological research promising great advancements for biodiversity and climate research. A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types. 
2.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
4.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@forsman15].

<!-- more refs:  Reich et al., 1997; Wright et al., 2004; Chave et al., 2009; Schrodt et al., 2015-->

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].   
On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel16; @lajoie15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].  
In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level. The analysis presented here is only a first step to disentangle within- and between-species variability. [...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment. [...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition. [...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role. The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research. [...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available,

- intra-specific variation and evolutionary trade-offs
- global variation
- community functional pattern, comparative study across continents
- big data challenges
-->

## Overview of trait-data initiatives

Despite these obvious benefit and grand prospects, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous. 
Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme. Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.

Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g. including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism --> 
<!-- Nadja: This is of course the major achievement of TRY, but I think we don't have to mention it because this not our immediate goal with the trait standard. The important bit for us is that we provide tools to harmonize datasets. -->

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level in community-aggregated traits [@krause14a; @fierer14; CAT, @fierer12] derived from metagenomic analysis. <!-- databases or datasets on microbial traits? -->

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @sokol11, @price13 <!-- Nadja: oldest relevant pub. in Dryad-->]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

@@

## Need for ontologies to enable data standardisation

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15]. However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits comply with this demand for open data. While it seems clear that factual data, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@ref], there is still no broad consensus achieved in the scientific community whether such data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.  <!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. legal standpoint in most countries: descriptive data are not copyright protected. Publicly fundet research must provide products. is still debated by the scientific community. -->

Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation. Although open for manual download, data on those platforms might be stored in proprietary (e.g. .xlsx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious or impossible. Since the platforms are open for any type of scientific data, no data standards are forced upon the uploaded material. 
Wheter open or not, the publication of original data is increasingly required by journals to warrant reproducibility of science. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.

Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczorek12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities. The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data at file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standads. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

# Proposing an ecological traitdata standard

With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

We build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12]. We further expand the definitions of these terms  and add new key terms for the use case of trait data, particularly considering the work of @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

## Common structure of trait datasets

@@

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'. 
To comply with the ETS, datasets must be reformatted to the longtable format (Fig. 2*c*).

@@

## Common terms for trait data

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above. To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers. 
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).

Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).

The online reference glossary of terms is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the ETS as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.  
Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically via an API (@ref). Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

<!--
Table 1: Minimal set of terms for the traitdata standard. valueType can be of type character, factor, boolean or numerical. vocabulary is only needed for factorial trait data and defines the factor levels which are allowed. The identifier is a unique and globally valid URI.

| Term | valueType | vocabulary | Identifier | Definition | 
|------|-----------|------------|------------|------------|
|scientificName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificname |Original character string provided as species name by the data owner (kept for reference only)| 
|taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected. Each entry should link to a published reference species list/taxonomy. Avoid synonyms.|
|taxonRank | factor | domain; kingdom; subkingdom; superhpylum [...] |http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonrank | The taxonomic rank of the most specific name in the scientificName. Recommended best practice is to use a controlled vocabulary.|
|kingdom | factor | |http://fdschneider.de/bexis_traits/traitdatastandard.html#kingdom | The full scientific name of the kingdom in which the taxon is classified.| 
|traitName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#traitname |Name of the trait as used by the data provider (kept for compatibility on provider side).|
|traitID|factor|see traitlist|http://fdschneider.de/bexis_traits/traitdatastandard.html#traitid ||
|traitValue|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvalue |Measured raw value or factor level for this species trait as measured and provided by the author.|
|traitUnit|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunit |Reports the unit that the author's raw data were measured in, if applicable (only for numeric values).|
|scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive name of the trait reported. This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
|traitValueStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvaluestd |Standardised measured value or factor level for this species trait. Numerical values must use the correct unit. Factor levels must be compliant with the thesaurus of traits.|
|traitUnitStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunitstd |The units associated with the measurementValue. The expected standard unit for this trait as defined in the traitlist/ thesaurus.|
|measurementID|	character||	http://fdschneider.de/bexis_traits/traitdatastandard.html#measurementid |A unique, dataset-specific identifier. Links multi-value trait measurements, e.g. x-y-z coordinates of a morphometric landmark, biochemical compound quantities for different chainlengths. In this case, the trait names must specifiy the sub-measurement, e.g. "landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurrenceid|A character string user defined by the data author. It is recommended best practice to use a globally unique identifier, if available, e.g. a museum ID referring to the individual specimen from which the data were measured. Usually does not apply for literature data.|
|warnings|character|	|	|Warnings from the R-script will be stored here, e.g. regarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table. User defined warnings and flags can be added as well, e.g. 'NOTUSE'.|

<!-- Nadja: Is this the kind of table you had in mind? -->

@@

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.

Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`). For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods). This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms. (more detail on semantic web ontologies can be found in the online reference for the ETS at @refs...).

@@

The ETS provides two extensions of the namespace that should be used to describe these datasets:

@@

For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf).

The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

1.	The assessment and analysis of functional traits has been established in all domains of ecological research, promising great advancements for biodiversity and climate research. A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types. 
2.	To harmonize and standardize datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
4.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07]. The scope of trait-based research spans from the evolutionary basis of individual-level properties to global patterns of biodiversity and ecosystem functioning.

Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parameterize ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. 
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97]. Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].

In recent years, functional ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. 
However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. 
Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, DataDryad). Given this prospect, there will be no lack of data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold struture and vocabulary for describing trait-datasets compatible with existing ontologies. (5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.

With an easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold  trait data standardisation.

# 1. Need for data standardisation

Despite the obvious benefit and grand prospects of functional approaches, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.

Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These compiled data are either being published as static data tables or in the form of sophisticated web-services (see below for a review of current initiatives). Services attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, but more importantly, provide platforms for consensus building in methods. Furthermore, database services offer access control and data usage policies, to claim the rights of the original data providers, while simplifying data queries for synthesis researchers.

Nonetheless, open-access publications in low-threshold data repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation. Although open for manual download, data on low-threshold data platforms might be stored in proprietary (e.g. .xlsx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious or impossible.

On the other hand, not all centralised databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories. 
Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit. 
Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship. 
Low-threshold data repositories assume public domain and Creative Commons Attribution licenses compliant with the demands of journals and research funding.

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
To cope with this trend towards distributed data sources, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. These services acquire public data from other sources and simplify the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

# 2. Overview of trait-data initiatives

In this section, we list two kinds of initiatives: the first aims at the harmonization of trait data into trait databases; the second develops trait ontologies and data standards. Often ontologies are being developed to serve a particular database project and thus are being developed in concert, as in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].

## Databases and Datasets

Plant flower, leaf and stem traits have been measured for decades, and were compiled into comprehensive databases and published as static datasets, e.g. the LEDA traitbase for the Northwest-European flora [@kleyer08]. 
Regarding webservices, most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11]. Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed.

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12; @ulrich08] derived from metagenomic analysis [@torsvik02; @langille13]. Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12].

In the animal kingdom, n unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17; @sokol11; @price13]. In the past two decades, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
Specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). 
Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@moretti17].

A general database for major physiological and life-history traits of all taxa is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].

@@

## Thesauri and Ontologies

A glossary or thesaurus of terms would be primarily a listing of definitions. Expanding on this, ontologies are linking individual terms with each other by formally defining the relationship between terms, to the ends of enabling a computational interpretation of data.

Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait standards, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measureing. For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.

While trait thesauri define traits for focal groups of organisms, they are not specifing the  formats or structures in which trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities. 
The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

# 3. Common structure of trait datasets

@@

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig. 2*c*).

@@

# 4. Proposing an ecological traitdata standard

With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices.

Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits (see below for further detail).

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the data standard as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.  
Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

## core structure for trait data

As described above, longtable formats are the most predictable and compatible format for complex datasets.

If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).

If taxon information and metadata need to be included, datasets may apply the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core Archives [DwCA;].

<!--- -TO DO ---->

## vocabulary for trait data

In the subsequent sections we propose a set of terms for composing a trait dataset, starting with the columns required for our minimal definition of 'core' trait-datasets proposed above. To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).

@@

Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`). For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms [@garnier17].

@@

The data standard provides two extensions of the namespace that should be used to describe these datasets:

@@

For a detailed description on how to use the package see the package documentation. 
The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17]. 
To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.  
We provide a script (see online appendix and R-package vignette <!-- add URL-->) to demonstrate the harmonization of very different data sources into our data standard.
